\section{Introduction}
\label{sec:introduction}

Deep learning on molecular sequences, from property prediction of drug-like molecules \citep{chithrananda2020chemberta,ross2022molformer} to protein language modeling \citep{lin2023esm2} and 3D molecular representations \citep{zhou2023unimol}, relies almost entirely on transformer architectures borrowed from natural language processing with minimal modification. This implicitly assumes that molecules and language share the same computational requirements. Yet molecular sequences differ from natural language in fundamental ways: SMILES strings use fewer than 50 characters, protein sequences use 20 amino acid tokens, and both are far shorter than typical NLP contexts. Whether these differences warrant distinct architectural designs has not been systematically investigated.

Neural architecture search (NAS) offers a principled way to discover domain-specific designs \citep{elsken2019nas}, but traditional NAS operates over hand-designed search spaces that may miss important structural innovations. Recent work has shown that language models can serve as program search agents, proposing open-ended code modifications that go beyond predefined search spaces \citep{chen2023evoprompting,romeraparedes2024funsearch,hu2025lmsearcher}. The autoresearch framework \citep{karpathy2026autoresearch} demonstrated that such an agent can iteratively improve a transformer training script for NLP through autonomous code modifications. However, this line of work has been limited to natural language, and no prior work has decomposed how much of the improvement comes from architectural changes versus hyperparameter tuning.

This paper makes three contributions:

\begin{enumerate}[nosep]
\item \textbf{A controlled decomposition of architecture search vs.\ hyperparameter tuning.} We design a 4-condition experiment (autonomous agent, random NAS, hyperparameter-only agent, fixed default) that cleanly separates the contributions of architecture search and hyperparameter tuning across three molecular and language domains (SMILES, protein, NLP), totaling 3{,}106 experiments.

\item \textbf{Evidence that the value of architecture search is domain-dependent.} On NLP, architecture search contributes 81\% of total improvement over the baseline ($p = 0.008$); on SMILES, hyperparameter tuning alone captures 151\% of the improvement ($p = 0.001$) and architecture search is counterproductive. For proteins, margins are too small for either component to reach significance.

\item \textbf{A surprising universality finding.} While agent-discovered architectures cluster by domain ($p = 0.004$), all 41 innovations transfer across domains with $<$1\% degradation ($p = 2 \times 10^{-19}$ against the predicted 35\% universal rate). The clustering reflects search path dependence, not fundamental domain requirements at this scale.
\end{enumerate}

These findings yield a practical decision framework: for molecular domains with small vocabularies and short sequences (like SMILES), teams should constrain their search to hyperparameters only; for domains with large vocabularies and long sequences (like NLP), full architecture search accounts for the majority of gains. We release the complete framework, all 3{,}106 experiment logs, and the agent prompts as open-source resources.
