\section{Practical Implications}
\label{sec:practical}

The decomposition results translate directly into compute budget allocation guidance.

\textbf{When to use architecture search vs.\ HP tuning.}
Table~\ref{tab:decision} summarizes a decision framework based on observable domain characteristics. The key predictors are vocabulary size and sequence length: domains with small vocabularies ($<$100 tokens) and short sequences ($<$500 tokens) resemble SMILES, where HP tuning alone suffices; domains with large vocabularies ($>$1K tokens) and long sequences ($>$1K tokens) resemble NLP, where architecture search contributes the majority of improvement.

\begin{table}[t]
\centering
\caption{Decision framework: recommended optimization strategy by domain characteristics. Based on the SMILES/protein/NLP decomposition results (Table~\ref{tab:decomposition}).}
\label{tab:decision}
\small
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Vocab size} & \textbf{Seq length} & \textbf{Resembles} & \textbf{Recommendation} \\
\midrule
$<$100 & $<$500 & SMILES & HP tuning only \\
$<$100 & 500--1{,}000 & Protein & Either; thin margins \\
$>$1{,}000 & $>$1{,}000 & NLP & Full architecture search \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Cost-benefit analysis.}
Each search condition costs approximately \$3.50--5.00 per 100 experiments (5 min $\times$ 100 $\times$ \$0.44/hr GPU $\approx$ \$3.67, plus ${\sim}$\$1.50 in LLM API costs for agent conditions). Random NAS avoids API costs entirely. For SMILES-like domains, the \$5 HP-only agent achieves the best result (mean best 0.581 bpb); for NLP-like domains, the \$5 full agent achieves the best result (mean best 1.123 bpb). Practitioners working with molecular data can run a full HP search overnight on a single GPU for under \$10.

\textbf{Transferable innovations.}
Four universally beneficial innovations emerged across all domains and can be applied to any transformer architecture: (1)~grouped query attention ($n_\text{kv\_head} = 1$; 5:1 KV compression with negligible quality loss), (2)~gated MLPs (SwiGLU/GeGLU replacing ReluSquared), (3)~learned per-layer residual scaling, and (4)~value embeddings on every layer rather than alternating layers. These transferred across all domain pairs with $<$1\% degradation.
