\section{Related Work}
\label{sec:related_work}

\textbf{LLM-guided program search.}
The use of LLMs as search agents for code-level optimization has emerged as an alternative to traditional NAS. EvoPrompting \citep{chen2023evoprompting} combines LLM code generation with evolutionary search for neural architecture design, while FunSearch \citep{romeraparedes2024funsearch} uses LLMs to discover novel mathematical programs. LM-Searcher \citep{hu2025lmsearcher} applies LLM-based NAS across multiple architecture domains. Self-Refine \citep{madaan2023selfrefine} provides a general framework for iterative LLM refinement, and IMPROVE \citep{xue2025improve} extends this to ML code optimization. The AI Scientist \citep{lu2024aiscientist} pursues fully autonomous scientific discovery including architecture design. ELM \citep{lehman2022elm} and LLMatic \citep{nasir2023llmatic} explore evolution-through-LLMs and quality-diversity optimization respectively. We differ in two respects: (1) we apply LLM-guided search to molecular domains, not just NLP, and (2) we introduce a 4-condition design that decomposes architecture search from HP tuning, a distinction absent from prior work.

\textbf{Autoresearch.}
Our framework builds on the autoresearch system \citep{karpathy2026autoresearch}, which demonstrated that an LLM agent can iteratively improve an NLP training script through autonomous code modifications. We extend autoresearch to molecular domains and, critically, add three baseline conditions that enable a controlled decomposition. The original autoresearch work does not separate architectural improvements from hyperparameter tuning, making it impossible to assess the value of architecture search per se.

\textbf{Neural architecture search.}
Traditional NAS methods operate over discrete \citep{elsken2019nas} or continuous search spaces with fixed parameterizations. Random search \citep{bergstra2012random} remains a strong baseline, and methods like BOHB \citep{falkner2018bohb} combine Bayesian optimization with bandit-based early stopping. Our random NAS baseline follows this tradition: uniform sampling from a discrete architectural space. The key distinction of LLM-guided search is that it operates on source code directly, enabling open-ended modifications (e.g., introducing gated MLPs or changing residual connections) that lie outside any predefined search space.

\textbf{Molecular transformers.}
Transformer-based models for molecular data have proliferated in recent years. ChemBERTa \citep{chithrananda2020chemberta} and Chemformer \citep{irwin2022chemformer} apply BERT- and encoder-decoder architectures to SMILES strings. MoLFormer \citep{ross2022molformer} introduces linear attention for molecular representations. ESM-2 \citep{lin2023esm2} trains protein language models at billion-parameter scale. Uni-Mol \citep{zhou2023unimol} incorporates 3D structural information. These models all adopt NLP architectures with minimal domain-specific adaptation. Our work provides the first systematic evidence for \emph{when} such adaptation matters: architecture search adds substantial value for NLP-like domains but not for SMILES-like domains at the ${\sim}$10M parameter scale.

\textbf{Scaling laws and compute-optimal training.}
Scaling laws \citep{kaplan2020scaling,hoffmann2022chinchilla} inform the relationship between model size, data, and compute. Our experimental design operates in a deliberately small-scale regime (${\sim}$8.6M parameters, 5-minute training) to maximize the number of architecture evaluations within a fixed budget. This mirrors the proxy-based evaluation common in NAS, where short training runs approximate longer training \citep{elsken2019nas}. We validate our 5-minute proxy against 2-hour training (Spearman $\rho = 0.54$) and discuss its limitations.

\textbf{Grouped query attention and gated MLPs.}
Two of the universally beneficial innovations our agent rediscovered, grouped query attention (GQA) \citep{ainslie2023gqa,touvron2023llama2} and gated linear units \citep{dauphin2017glu,shazeer2020glu}, are well-established in the NLP literature. That the agent converges on these independently, without explicit prompting, supports the finding that effective transformer innovations are largely domain-agnostic at this scale.
