\section*{Reproducibility Statement}
\label{sec:reproducibility}

\textbf{Code and data.} The complete framework (training script, agent prompts, data preparation pipelines, analysis scripts, and all 3{,}106 experiment logs including intermediate \texttt{train.py} versions and code diffs) is available at \url{https://github.com/ewijaya/autoresearch-mol}. All datasets are publicly available: ZINC-250K, UniRef50, and FineWeb-Edu.

\textbf{Experimental setup.} All experiments run on a single NVIDIA A10G GPU (24\,GB VRAM) with a fixed 5-minute training budget. Random seeds are fixed (\texttt{torch.manual\_seed(42)}). The random NAS baseline uses deterministic seeds per replicate. Model architecture, optimizer, and training hyperparameters are fully specified in Section~\ref{sec:methodology} and in the released \texttt{train.py}.

\textbf{Statistical analysis.} The analysis script (\texttt{scripts/analyze\_phase2.py}) reads only on-disk results and produces all figures, tables, and hypothesis tests reported in this paper. Bootstrap CIs use 10{,}000 resamples. Permutation tests use 10{,}000 permutations. All $p$-values are reported with both raw and Holm--Bonferroni adjusted values.

\textbf{Limitations on exact reproducibility.} The LLM agent (OpenAI Codex / GPT-5.4) is accessed via API and is non-deterministic: individual agent runs are not exactly reproducible. We address this by running 3--5 independent replicates per condition and reporting aggregate statistics with confidence intervals. The random NAS and fixed default conditions are fully deterministic.
